---
title: "Aluminum Data"
author: Hong Zhang
date: 2026-09-24
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle
<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

### __q1__ Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>%
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    cols = c(-thick, -alloy)
  ) %>%
  mutate(angle = as.integer(angle)) %>%
  filter(E > 0, nu > 0)

df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA
<!-- -------------------------------------------------- -->

## Initial checks
<!-- ------------------------- -->

### __q2__ Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
df_stang_long %>%
  summary()

df_stang_long %>%
  count(alloy)

df_stang_long %>%
  count(angle)

df_stang_long %>%
  count(thick)

# number of plates at each thickness (each row of the raw data is one plate)
df_stang %>%
  count(thick)

# repeated tests under the same conditions
df_stang_long %>%
  filter(thick == 0.022, angle == 45)
```

**Observations**:

- Is there "one true value" for the material properties of Aluminum?
  - No, there is not one true value for the material properties of aluminum. Even for a single alloy, the summary shows E ranges from 9900 to 10700 and nu ranges from 0.310 to 0.331 across the 26 measurements. Repeated tests under the same conditions also disagree. Filtering to the 0.022 thick plates tested at 45 degrees gives two measurements with E values of 10700 and 10500.
- How many aluminum alloys are in this dataset? How do you know?
  - There is one aluminum alloy in this dataset, al_24st. count(alloy) returns a single row containing all 26 measurements.
- What angles were tested?
  - The angles tested were 0, 45, and 90 degrees. There are 9 measurements at 0 degrees, 8 at 45 degrees, and 9 at 90 degrees. The 45 degree group has one fewer because the last row of the raw df_stang table has -1 for E_45 and nu_45, which was removed as invalid in q1.
- What thicknesses were tested?
  - The thicknesses tested were 0.022, 0.032, 0.064, and 0.081. There are 6 measurements each at 0.022, 0.032, and 0.064, and 8 measurements at 0.081. Counting the rows of the raw df_stang table shows that 3 plates were tested at 0.081 and 2 plates at each other thickness, which is why 0.081 has more measurements.
- Does the angle relative to the rolling direction affect the elasticity E? (own question)

## Visualize
<!-- ------------------------- -->

### __q3__ Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
df_stang_long %>%
  ggplot(aes(angle, E)) +
  geom_point(size = 3) +
  facet_wrap(~ thick)
```

**Observations**:

- Based on this data, angle does not appear to have a consistent effect on E. In the 0.022 panel, every measurement at every angle falls between 10500 and 10700. In the 0.064 and 0.081 panels, the 0 degree points are slightly higher than the points at the other angles, but in the 0.032 panel the 45 degree points are highest instead. No angle is highest across all four thicknesses.
- Where the points at one angle are higher, the difference is 200 or less, which is no bigger than the difference between two plates tested at the same angle. For example, the two 0.022 plates tested at 45 degrees gave 10500 and 10700. Any effect of angle is too small to separate from the plate to plate variation in this data.
- This answer is not conclusive. There are only two or three measurements for each thickness and angle, and every E value is recorded to the nearest 100, so an effect smaller than about 100 could not be detected. Answering the question would require more plates tested at each thickness and angle, with E recorded more precisely than the nearest 100.
- Some panels show fewer dots than measurements because identical values overlap exactly. For example, the two 0.022 plates at 0 degrees both measured 10600.

### __q4__ Consider the following statement:

> "A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

- Does this graph support or contradict the claim above?
  - The graph partly contradicts the claim that E is an intensive property. The 0.081 thick plates all have E between 9900 and 10100, while every plate at the other three thicknesses has E of 10300 or higher. If E did not depend on the amount of material, we would not expect the points for the thickest plates to be clearly separate from the rest. However, E shows no trend across the other three thicknesses. The 0.022, 0.032, and 0.064 plates overlap between 10300 and 10700, and the 0.064 plates tend to have higher E than the thinner 0.032 plates. For those three thicknesses, E looks about the same regardless of thickness, which supports the claim.

- Is this evidence *conclusive* one way or another?
  - No, this evidence is not conclusive. The difference comes from a single thickness, 0.081, rather than a steady change across all thicknesses. The lower E could be caused by thickness or by something about those specific plates, such as coming from a different batch of material. There are also only two or three plates at each thickness, all of the same alloy. To decide, we would need plates at more thicknesses, especially between 0.064 and 0.081 and above 0.081, to see whether E keeps dropping as thickness increases. We would also need 0.081 thick plates from more than one batch of material to see whether the low E comes from the thickness or from those particular plates.

# References
<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[https://en.wikipedia.org/wiki/List_of_materials_properties]
